---
title: "class15"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The data for for hands-on session comes from GEO entry: GSE37704, which is associated with the following publication:

>Trapnell C, Hendrickson DG, Sauvageau M, Goff L et al. "Differential analysis of gene regulation at transcript resolution with RNA-seq". Nat Biotechnol 2013 Jan;31(1):46-53. PMID: 23222703

## Import count and metadata

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```

remember that we need the countData and colData files to match up so we will need to remove that odd first column in countData namely countData$length.

```{r}
# Note we need to remove the odd first $length col
countData <- as.matrix(countData[,-1])
head(countData)
```

Double check that the colnmes in countData match the id values in the metadata file.

```{r}
colnames(countData)
rownames(colData)

all( colnames(countData) == rownames(colData) )
```

```{r}
# Test how all() function works
all( c(T, T, T) )
all( c(T, F, T) )
```

## Remove zero count genes

We want to remove genes that have 0 count values in all experiments (i.e. rows that have 0 across all cols)

```{r}
# Filter out
countData = as.matrix(countData[rowSums(countData) != 0, ])
head(countData)
```

# DESeq analysis

```{r}
library(DESeq2)
```

```{r}
# Setup the object with our data in the way DESeq wants it
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
# Run the analysis
dds = DESeq(dds)
```

Get our results

```{r}
res = results(dds)

res
```

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

```{r}
# Make a color vector for all genes
mycols <- rep("grey", length(res$padj))

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2] <- "red"

# Color blue those with adjusted p-value less than 0.01 and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2)
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

# Adding gene zymbols and entrez ids

```{r eval = FALSE}
BiocManager::install("AnnotationDbi")
BiocManager::install("org.Hs.eg.db")
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

# We can translate betwen al the following databse ID systems
columns(org.Hs.eg.db)
```

```{r}
res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(countData), # where are my IDs
                    keytype="ENSEMBL",    # what format are my IDs
                    column="SYMBOL",      # the new format I want
                    multiVals="first")
                    
res
```

```{r}
res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(countData),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res
```

```{r}
res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

res
```

# Pathway analysis

```{r eval = FALSE}
BiocManager::install( c("pathview", "gage", "gageData") )
```

```{r}
library(pathview)
library(gage)
library(gageData)
```

```{r}
data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

The main gage() function requires a named vector of fold changes, where the names of the values are the Entrez gene IDs.

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

now lets look at the object returned from gage()

```{r}
attributes(keggres)
```

```{r}
# Look at the first few down (less) pathways
head(keggres$less)
```

Now, let's try out the pathview() function from the pathview package to make a pathway plot with our RNA-Seq expression results shown in color.

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

This downloads the pathway figure data from KEGG and adds our results to it. 

You can play with the other input arguments to pathview() to change the display in various ways including generating a PDF graph. 

```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

Now, let's process our results a bit more to automagicaly pull out the top 5 upregulated pathways, then further process that just to get the pathway IDs needed by the **pathview()** function. We'll use these KEGG pathway IDs for pathview plotting below.

```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids <- substr(keggrespathways, start=1, stop=8)
keggresids
```

Finally, lets pass these IDs in keggresids to the pathview() function to draw plots for all the top 5 pathways.

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species = "hsa")
```

Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?

```{r}
keggrespathways.down <- rownames(keggres$less)[1:5]
keggresids.down <- substr(keggrespathways.down, start = 1, stop = 8)
pathview(gene.data = foldchanges, pathway.id = keggresids.down)

```


# Gene Ontology (GO)

We can also do a similar procedure with gene ontology. Similar to above, go.sets.hs has all GO terms. go.subs.hs is a named list containing indexes for the BP, CC, and MF ontologies. Let’s focus on BP (a.k.a Biological Process) here.

KEGG focuses on metabolic pathway/biochem part, whereas the GO focuses on the biological process

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir = TRUE)

lapply(gobpres, head)
```

